---
title: "Introductory Computer Programming Assignment 1"
author: "Soumya Mukherjee"
date: "31 October 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, echo=FALSE}
library(Rcpp)
library(lattice)
```
## 1. Recursive Version of Insertion Sort :

* The basic problem:

	* Input: A sequence of numbers $(a_1, a_2, ..., a_n)$

	* Desired output: A permutation of the input, $(b_1, b_2, ..., b_n)$
      such that $b_1 \leq b_2 \leq ... \leq b_n$

* The $a_i$-s are known as _keys_.

* For sorting, we usually need a simple data structure known as an
_array_:

	* An array $A[1,...,n]$ of length $n$ is a sequence of length $n$.

	* The $i$-th element of an array $A$ is denoted by $A[i]$

	* Each $A[i]$ acts as a _variable_, that is, we can assign values
      to it, and query its current value

	* The sub-array with indices $i$ to $j$ (inclusive) is often
      indicated by $A[i, ..., j]$

### Insertion sort:

* _Insertion sort_ is a simple and intuitive sorting algorithm

* Basic idea of Insertion sort(Iterative version):

	1. If the array is of length 1 it is trivially sorted.
	
	2. Else choose the first element of the input array.

	3. Compare the second element of the array with the first. If the first element is greater than the second, then swap them else leave their positions in the array unchanged.

	4. Having sorted the first two elements, insert the third element into its correct sorted position considering only the first three elements of the array now.

	5. Continue in this manner until the last element of the array is inserted into its correct sorted position in the array.
	
* Basic idea of Insertion sort (Recursive version):

	1. If the array is of length 1 it is trivially sorted.

	2. Else divide the array into two parts - the first part containing the first n-1 elements of the array in the given order and the second part consisting of the $n-th$ element of the array, n being the length of the array.

	3. If the first part is of length 1 then it is trivially sorted.Then compare the second part of the array i.e. the last element of the array, with it and swap them if they are not in nondecreasing order.
	
	4. If the first part has length greater than 1, apply this algorithm on these n-1 elements to sort them i.e. use this algorithm recursively. After they have been sorted, insert the last element of the original n-element input array into its correct sorted position.
	
###Pseudo Code for recursive version of insertion sort:

Here the input is an already-constructed array `A` along with the no. of elements of the array starting from the first that we want to sort, given by 'n'. If we want to sort the whole array then 'n' is the length of the array.

####Pseudo Code for the algorithm:

\

\

`recur.ins.sort(A, n)`

| if ( n < 2 ) return (A)
| else {
|     A = `recur.ins.sort(A, n-1)`
|     key = A[n] 
|     j = n - 1
|     while (j > 0 and A[j] > key)
|     {
|         A[j+1] = A[j]
|         j = j - 1
|     }
|     A[j+1] = key
|     return (A)
| }

##2. Correctness of the algorithm:

We shall prove the correctness of the algorithm by recursion invariant.

#### Statement

>  At the beginning of each recursion to sort the fitst n-j+1 elements of the array (for any particular value of $j$ between 1 and n-1),
>  The first $n-j$ elements in $A[1,...,n-j]$ are the same as the
>  first $n-j$ elements originally in the array, but they are now in sorted order,
>  where n is the length of the input array.

#### Initialisation 

* For $j = n-1$ , the Array $A[1,2,..., n-j]$ has exactly one element which is trivially sorted.
* Hence the above statement holds for $j = n-1$

#### Maintenace 

* Suppose the recursion invariant is true for index $j$.
* At the beginning of the recursion to sort the first n-j+1 elements of the array (for any particular value of $j$), $A[1,2,...,n-j]$ is sorted.
* While loop within each recursion works by 
    * comparing $key = A[n-j+1]$ with $A[n-j],A[n-j-1],...,A[1]$ (following this order)
    * moving them one position to the right , until the correct position of $key$ is found
* Evidently, this while loop must terminate in atmost $n-j$ steps.
* At the end $A[1,2,...,n-j+1]$ is a sorted version of original array $A[1,2,..,n-j+1]$
* Thus, the recursion invariant is true for index $j-1$.

#### Termination 

* Each call of the function decreases the length of sub-array $A[1,...,n-j]$ by $1$ ,
* The function stops recursive calls when the length of sub-array $A[1,...,n-j] = 1$ i.e. when $j=1$.   
* Therefore the recursion terminates.

    * Hence the algorithm is correct.

##3. Average case Runtime:

* When inserting $A[n]$ in its correct position , it assumes that $A[1,2,..,n-1]$ is already sorted for any value of $n$.
* So, in the best case scenario the while loop will have to compare $A[j] < key = A[n]$ only once, whereas in the worst case scenario the while will loop have to do the same $n-1$ times.
* So for randomised input, on an average the while loop will have to do the comparison $\frac{n}{2}$ times (As, the number of comparisons has distribution $Uniform${1,2,...,n-1} )
* Hence the recursive relation for Average case runtime $T(n)$ is given by:

$$T(n) =
\begin{cases}
  O(1) & n = 1 \\
  T(n-1)+ O(\frac{n}{2}) & n > 1
\end{cases}$$

* Therefore Average case runtime is given by:

\begin{equation}
\begin{aligned}
T(n) &= T(n-1) + O(\frac{n}{2})\\
&= O(\frac{n}{2}) + O(\frac{n-1}{2}) + \dots + O(\frac{2}{2}) + T(1)\\
&= O(\frac{n(n+1)}{4})\\
&= O(n^2)
\end{aligned}
\end{equation}

##4. Implementations of the algorithm in R and Rcpp:

###Insertion Sort in R:

####1. Iterative Version of Insertion Sort:

* The iterative insertion sort is implemented in R as follows:


```{r, echo=TRUE}
iter.ins.sort = function(A){
  if (length(A) < 2) return(A)
  for (j in 2:length(A)) {
    key = A[j]
    i = j-1
    while (i > 0 && A[i] > key) {
      A[i+1] = A[i]
      i = i-1
    }
    A[i+1] = key
  }
  return(A)
}
```


####2. Recursive Version of Insertion Sort:


* Recursive version of insertion sort is implemented in R as follows :


```{r, echo=TRUE}
recur.ins.sort = function (A, n = length(A) ){
  if ( n < 2 ) return(A)
  else{
    A = recur.ins.sort(A, n-1)
    key = A[n]
    j = n-1
    while (j > 0 && A[j] > key ) {
      A[j+1] = A[j]
      j = j-1
    }
    A[j+1] = key
    return(A)
  }
}
```

### Insertion Sort using Rcpp:

####1. Iterative Version of Insertion Sort using Rcpp

```{Rcpp, cache=TRUE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector iter_ins_sort_rcpp(NumericVector A)
{
	int i, n = A.size();
	double key;
  for (int j = 1; j < n; j++)
  {
      key = A[j];
      i = j - 1;
      while (i > -1 && A[i] > key)
      {
          A[i+1] = A[i];
          i = i - 1;
		  }
      A[i+1] = key;
	}
	return A;
}
```


####2. Recursive Version of Insertion Sort using Rcpp


```{Rcpp, cache=TRUE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector recur_ins_sort_Rcpp(NumericVector A, int n )
{
	int j;
	double key;
  if (n < 1)
  {
    return (A);
  }
  else
  {
    A = recur_ins_sort_Rcpp( A, n-1 );
    key = A[n-1];
    j = n-2;
    while(j > -1 && A[j] > key)
    {
      A[j+1] = A[j];
      j = j-1;
    }
    A[j+1] = key;
  }
  return (A);
}
```

 
### Some sample implementations of these programs and the inbuilt sort function in R: 


```{r,echo=TRUE}
A = sample(1:1000, 50, replace = TRUE)
A
sort(A)
A = sample(1:1000, 50, replace = TRUE)
A
iter.ins.sort(A)
A = sample(1:1000, 50, replace = TRUE)
A
recur.ins.sort(A, length(A))
A <- sample(1:1000, 50, replace = TRUE)
A
iter_ins_sort_rcpp(A)
A <- sample(1:1000, 50, replace = TRUE)
A
recur_ins_sort_Rcpp(A, length(A))
```


##5. Empirical Runtime Comparison:

* We define a function `timeSort` which will calculate the average runtime required to sort an array.
* For the recursive version of the programs we define two other  functions namely `sin.recur.ins.r` and `sin.recur.ins.rcpp` for R and Rcpp programs respectively , so that these functions would require only asingle argument instead of two arguments and consequently we will be able to run `timeSort` on them, otherwise obtsining and compaing empirical runtimes takes a very long time using for loops or some similar variants.
* With the large length of input array (above 2300) the recursive version implementations in both R and Rcpp , get teminated abruptly due to "Stack error" or "deep nested infinite recursions"  .

```{r}
timeSort <- function(size, nrep , sort.fun )
{
    x <- replicate(nrep, runif(size), simplify = FALSE)
    system.time(lapply(x, sort.fun))["elapsed"] / nrep
}
n <- seq(100, 2300, by = 50)
sin.recur.ins.r <- function(A){
  n = length(A)
  recur.ins.sort(A,n)
}

sin.recur.ins.rcpp <- function(A){
  n = length(A)
  recur_ins_sort_Rcpp(A,n)
}
set.seed(0)
t_iter_ins_r <- sapply(n, timeSort, nrep = 5, 
                                       sort.fun = iter.ins.sort)
t_recur_ins_r <- sapply(n, timeSort, nrep = 5 , 
                                       sort.fun = sin.recur.ins.r)
t_iter_ins_rcpp <- sapply(n, timeSort, nrep = 5, 
                                          sort.fun = iter_ins_sort_rcpp)
t_recur_ins_rcpp <- sapply(n, timeSort, nrep = 5 , 
                                          sort.fun = sin.recur.ins.rcpp)
t_sort_r <- sapply(n, timeSort, nrep = 5, sort.fun = sort)

```

We plot the empirical runtimes of the different sorting algorithms against input size individually and again together on the same plot for comparing them.
```{r}
xyplot(t_iter_ins_r + t_recur_ins_r + t_iter_ins_rcpp + t_recur_ins_rcpp
       + t_sort_r ~ n , grid = TRUE , outer = TRUE, ylab = "time (seconds)",
       scales = list(y = "free", x = "free"),
       main="Individual Empirical Runtimes for different sorting algorithms")
```

\newpage

```{r}
xyplot(t_iter_ins_r + t_recur_ins_r + t_iter_ins_rcpp + t_recur_ins_rcpp
       + t_sort_r ~ n , grid = TRUE , ylab = "time (seconds)",
       scales = list(y = "free", x = "free"), auto.key = TRUE ,
       main="Comparison of Empirical Runtimes for different sorting algorithms")
```

### Note:

* Observations for each implementation individually:

    * Iterative version using R - The empirical runtime for this implementation appears to be monotonically increasing in a quadratic manner with respect to the increasing input array size , with runtime reaching a maximum of 0.20 seconds for input array size 2300. As far as we tested the program sorted correctly even upto input array sizes 5000.
    * Recursive version using R - The empirical runtime for this implementation appears to be monotonically increasing in a quadratic manner with respect to the increasing input array size , with runtime reaching a maximum of 0.25 seconds for input array size 2300. As noted due to stack error the program does not run for larger input arrays.
    * Iterative version using Rcpp - The empirical runtime for this implementation appears to be almost zero for majority of input array sizes, and for the rest the difference is negligible with respect to 0.This implementation correctly sorts arrays even upto sizes 5000 and larger.
    * Recursive version using Rcpp - The empirical runtime for this implementation appears to be almost zero for almost every input array size, and for the rest the difference is negligible with respect to 0. As noted due to deep nested recursion the program does not run for larger size input arrays than 2300.
    * Inbuilt sort function in R - The empirical runtime for this implementation appears to be almost zero for almost every input array size, and for the rest the difference is negligible with respect to 0. This implementation runs for every input array size we could think of.

* Observations regarding comparison of the implementations:

The empirical runtimes for the iterative and recursive versions of insertion sort using Rcpp along with the inbuilt sort function are almost zero for every input array size upto 2300, after which we cannot compare since the first two implementations do not work for larger input array sizes. Within this range of input array sizes The empirical runtimes of the iterative and recursive versions of insertion sort using R are significantly higher than the other three implementations and the runtimes for them are comparable among themselves upto around input sizes 1800 to 2000, both increasing monotonically in a quadratic manner with increase in input array size and for array sizes in the range 2000 to 2300, the iterative version using R performs better than its corresponding recursive version.  
  